---
title: "Setup"
author: "Joe Willage"
date: "March 23, 2016"
output: 
  html_document: 
    keep_md: yes
---

This is a first attempt at analyzing the NYC Street Tree Map data, specifically to find out which 
are the most tree-lined blocks.

```{r, warning = FALSE, message = FALSE}
library(ggmap)
library(knitr)
library(rjson)
library(dplyr)
```

```{r global_options, include = FALSE}
knitr::opts_chunk$set(fig.width = 7, fig.height = 7, fig.path='Figs/', 
                      warning = FALSE, message = FALSE, cache = TRUE)
```

Data is pulled from the NYC Street Tree Map site[^1]. The call takes in diagonal 
points and creates an MBB using those points as the southwest and northeast corners. The coordinates
below roughly describe a rectangle between the Manhattan bridge up to 8th and D. This is a 
reasonable area to start with while everything gets checked out and validated. 

```{r, eval = FALSE}
border.corner.sw.lat <- "40.71455760597046"
border.corner.sw.long <- "-73.99566650390625"
border.corner.ne.lat <- "40.723713744687295"
border.corner.ne.long <- "-73.97626876831055"

treemap.url <- paste("https://tree-map.nycgovparks.org/points",
              border.corner.sw.lat, border.corner.sw.long, 
              border.corner.ne.lat, border.corner.ne.long,
              "undefined", "1", "2000", sep = "/")
raw <- fromJSON(file = treemap.url)
```

The json returned includes the latitude and longitude for each tree in the range, as well as the 
tree species, trunk diameter, and a few other data points. At this point the data is trimmed down
from 1976 to a sample of 20 trees. These 20 will be used for validation. 

```{r, include = FALSE}
raw <- readRDS("rds/raw.rds")
```

```{r}
samp <- raw$item[1:20]
samp[[1]]
```


The ArcGIS API[^2] is used to reverse geocode each data point. It returns the closest address to 
each tree.  

```{r, eval = FALSE}
gis.url <- paste0("http://geocode.arcgis.com/arcgis/rest/services/World/GeocodeServer/",
                 "reverseGeocode?f=json&location=")
treeMap <- lapply(samp, function(i) {
  address <- fromJSON(file = paste0(gis.url, i$lng, ",", i$lat))
  c(i$id, i$lat, i$lng, i$stumpdiameter, address$address$Match_addr)
  })

treeMap <- data.frame(matrix(unlist(treeMap.raw), ncol = 5, byrow = TRUE), stringsAsFactors = FALSE)
treeMap$X5 <- gsub(", (New York|Knickerbocker), New York", "", treeMap$X5)
treeMap <- treeMap %>% separate(col = X5, into = c("address", "zip"), sep = ", ") %>% 
                    separate(col = address, into = c("number", "street"), extra = "merge")
names(treeMap)[1:4] <- c("id", "lat", "lon", "stumpDiam")
treeMap$stumpDiam <- as.numeric(treeMap$stumpDiam)
treeMap$number <- as.numeric(treeMap$number)
treeMap$lat <- as.numeric(treeMap$lat)
treeMap$lon <- as.numeric(treeMap$lon)
```

```{r, include = FALSE}
treeMap <- readRDS("rds/treeMap.sample.rds")
```

```{r}
head(treeMap)
```

The next thing to do is build a reference table consisting of all the blocks, which are defined 
as street segments between two adjacent intersecting streets. The Census geocoding service[^3] 
provides start and end addresses for a block. Each tree has it's address looked up in the block 
table. If there is no row whose address range includes the tree, a new block will be added via the
add block function. 

```{r eval = FALSE}
census.url <- "http://geocoding.geo.census.gov/geocoder/locations/address?"
# seed block table with correct data types
blocks <- data.frame(id = 1, start = 222, end = 276, street = "Elizabeth St", 
                     cross1 = "Prince St", cross1.lat = 40.722725, cross1.lon = -73.994156,
                     cross2 = "E Houston St", cross2.lat = 40.724434, cross2.lon = -73.99346, 
                     zip = "10012", count = 0, stringsAsFactors = FALSE)
treeMap$blockId <- NULL
addBlockErr <- NULL
for (i in 1 : nrow(treeMap)) {
  x <- inner_join(treeMap[i, ], blocks, by = c("street" = "street", "zip" = "zip"))
  if (nrow(x) > 0) {
    y <- x$start <= unique(x$number) & x$end >= unique(x$number)
    if (sum(y)) {
      # Taking only first match, not differentiating between sides of street
      blocks[x[which(y)[1], "id.y"], "count"] <- blocks[x[which(y)[1], "id.y"], "count"] + 1
      treeMap[i, "blockId"] <-  x[which(y)[1], "id.y"]
    } else {
      blocks <- rbind(blocks,  addBlock(treeMap[i, ]))
      treeMap[i, "blockId"] <- max(blocks$id)
    }
  }
  else{
    blocks <- rbind(blocks,  addBlock(treeMap[i, ]))
    treeMap[i, "blockId"] <- max(blocks$id)
  } 
}
```

```{r, eval = FALSE}
addBlock <- function(tree) {
  tryCatch({
    census.parms <- paste0("street=", tree$number, "+", 
                        gsub(" ", "+", tree$street), "&zip=", tree$zip, 
                        "&city=new+york&state=NY&benchmark=9&format=json")
    result <- fromJSON(file = paste0(census.url, census.parms))
    from <- as.numeric(result$result$addressMatches[[1]]$addressComponents$fromAddress)
    to <- as.numeric(result$result$addressMatches[[1]]$addressComponents$toAddress)
    # geocode addresses at each side of block
    from.parms <- paste0("street=", from, "+", 
                         gsub(" ", "+", tree$street), "&zip=", tree$zip, 
                         "&city=new+york&state=NY&benchmark=9&format=json")
    to.parms <- paste0("street=", to, "+", 
                       gsub(" ", "+", tree$street), "&zip=", tree$zip, 
                       "&city=new+york&state=NY&benchmark=9&format=json")
    fromCoord <- fromJSON(file = paste0(census.url, from.parms))
    toCoord <- fromJSON(file = paste0(census.url, to.parms))
    # lookup cross street
    fromInt <- fromJSON(file = paste0(gis.url, fromCoord$result$addressMatches[[1]]$coordinates$x, 
                                      ",", fromCoord$result$addressMatches[[1]]$coordinates$y,
                                      "&returnIntersection=true"))
    toInt <- fromJSON(file = paste0(gis.url, toCoord$result$addressMatches[[1]]$coordinates$x, 
                                    ",", toCoord$result$addressMatches[[1]]$coordinates$y,
                                    "&returnIntersection=true"))
    if (grepl(tree$street, fromInt$address$Address)) {
      cross1 <- gsub(tree$street, "", fromInt$address$Address)
      cross1 <- gsub(" & ", "", cross1)
    } else{
      warning("Intersection does not include original street")
    }
    if (grepl(tree$street, toInt$address$Address)) {
      cross2 <- gsub(tree$street, "", toInt$address$Address)
      cross2 <- gsub(" & ", "", cross2)
    } else{
      warning("Intersection does not include original street")
    }
    # order so multiple segment entries will match
    fromCross <- min(cross1, cross2)
    toCross <- max(cross1, cross2)
    # geocode intersection by names. More accurate point than start/end address
    gis.url.parms1 <- URLencode(paste0(tree$street, " and ", fromCross, ", nyc"))
    coords.cross1 <- paste0(gis.url.find, "&text=", gis.url.parms1)
    gis.url.parms2 <- URLencode(paste0(tree$street, " and ", toCross, ", nyc"))
    coords.cross2 <- paste0(gis.url.find, "&text=", gis.url.parms2)
    coords.cross1 <- fromJSON(file = coords.cross1)
    coords.cross2 <- fromJSON(file = coords.cross2)
    tmp <- data.frame(id = nrow(blocks) + 1, start = min(from, to), end = max(from, to), 
                      street = tree$street, cross1 = fromCross, 
                      cross1.lat = coords.cross1$locations[[1]]$feature$geometry$y,
                      cross1.lon = coords.cross1$locations[[1]]$feature$geometry$x,
                      cross2 = toCross,
                      cross2.lat = coords.cross2$locations[[1]]$feature$geometry$y,
                      cross2.lon = coords.cross2$locations[[1]]$feature$geometry$x,
                      zip = tree$zip, count = 1, stringsAsFactors = FALSE)
    tmp},
    error = function(err) {
      addBlockErr <<- c(addBlockErr, tree$id)
      tmp <- data.frame(id = nrow(blocks) + 1, start = min(from, to), end = max(from, to), 
                        street = tree$street, 
                        cross1 = NA, cross1.lat = NA,cross1.lon = NA,
                        cross2 = NA, cross2.lat = NA,cross2.lon = NA,
                        zip = tree$zip, count = 1, stringsAsFactors = FALSE)
      tmp
    }
  )
}
```

```{r, include = FALSE}
blocks <- readRDS("rds/blocks.sample.rds")
```

This is a good point to check the validity of all the calls.  

```{r sampleIntersections}
mymap <- get_map(location = "40.72017399459069,-73.98639034958494", zoom = 15, 
                 maptype = "toner-lines")
blocks$int1 = paste(blocks$street, "&", blocks$cross1)
blocks2 <- data.frame(int2 = paste(blocks$street, "&", blocks$cross2),
                      cross2.lat = blocks$cross2.lat, 
                      cross2.lon = blocks$cross2.lon)
g <-  ggmap(mymap) + 
      theme_nothing() +
      geom_point(aes(x = cross1.lon, y = cross1.lat), data = blocks, color = "red", size = 5) +
      geom_text(aes(label = int1, x = cross1.lon, y = cross1.lat), data = blocks, color = "blue",
                check_overlap  = TRUE) +
      geom_point(aes(x = cross2.lon, y = cross2.lat), data = blocks2, color = "red",   size = 5) +
      geom_text(aes(label = int2, x = cross2.lon, y = cross2.lat), data = blocks2, color = "blue",
                 check_overlap  = TRUE)
g
```

Everything is lining up correctly wrt intersections. A closer point-to-point inspection reveals
minute discrepancies. For instance tree 216757 is reverse geocoded as being next to 464 Grand, but
in reality is closer to 293 E Broadway, across the street. The tree will plot correctly on the map,
but it won't fall into the correct bucket when aggregating trees by block. This is an acceptable
margin of error at this point, but in the future, there are other options to explore reverse 
geocoding. 

The same process is performed on the original data set of 1976 trees. 

```{r, include = FALSE}
treeMap <- readRDS("rds/treeMap.rds")
blocks.les <- readRDS("rds/blocks.orig.rds")
```


```{r}
summary(blocks.les)
incomplete <- blocks.les[!complete.cases(blocks.les),]
sum(incomplete$count)
```

The resulting blocks table has `r nrow(blocks.les)` entries, with 
`r sum(!complete.cases(blocks.les))` NA's. Examining one of the NA's, 1 - 99 Willett
St (id == 179), the first intersection returns correctly at Willett and Delancey. However, the other
end of the block, 99 Willett, is incorrectly reverse geocoded up at Stanton St. Those incomplete 
blocks account for `r sum(incomplete$count)` trees, so it will be important to fill them in. To make 
the manual process easier, a subset of the `addBlock` function is broken out into the
`geocodeBlockEnds` function.

```{r}
gis.url.find <- "http://geocode.arcgis.com/arcgis/rest/services/World/GeocodeServer/find?f=json"
geocodeBlockEnds <- function(street, cross1, cross2){
  # given cross streets, add x and y coords
  
  # order so multiple segment entries will match
  fromCross <- min(cross1, cross2)
  toCross <- max(cross1, cross2)
  gis.url.parms1 <- URLencode(paste0(street, " and ", fromCross, ", nyc"))
  coords.cross1 <- paste0(gis.url.find, "&text=", gis.url.parms1)
  gis.url.parms2 <- URLencode(paste0(street, " and ", toCross, ", nyc"))
  coords.cross2 <- paste0(gis.url.find, "&text=", gis.url.parms2)
  coords.cross1 <- fromJSON(file = coords.cross1)
  coords.cross2 <- fromJSON(file = coords.cross2)
  l <- list(street = street, 
            cross1 = list(
              street = cross1,
              y = as.numeric(coords.cross1$locations[[1]]$feature$geometry$y),
              x = as.numeric(coords.cross1$locations[[1]]$feature$geometry$x)
            ),
            cross2 = list(
              street = cross2,
              y = as.numeric(coords.cross2$locations[[1]]$feature$geometry$y),
              x = as.numeric(coords.cross2$locations[[1]]$feature$geometry$x)
            )
  )
  l
}

head(incomplete)
```

```{r}
g <- geocodeBlockEnds("Baruch Pl", "Stanton St", "E Houston St")
blocks.les[blocks.les$id == 2, c(5:10)] <- c(g$cross1$street, g$cross1$y, g$cross2$x,
                                             g$cross2$street, g$cross2$y, g$cross2$x)
blocks.les[blocks.les$id == 2, ]
```

That assignment has to be repeated for each of the other 22 NAs, manually entering the correct 
cross streets. See [corrections file](https://github.com/jwillage/Trees/blob/master/corrections.R) 
for the full list of corrected segments. And some further error checking for bad segments...

```{r}
same <- blocks.les[blocks.les$cross1 == blocks.les$cross2, ]
nrow(same)
```

After correcting these manually (see corrections file), the previously bad intersections look as 
follows

```{r, include = FALSE}
blocks.les <- readRDS("rds/blocks.corrected.rds")
```


```{r correctedIntersections}
blocks.les$cross1.lat <- as.numeric(blocks.les$cross1.lat)
blocks.les$cross1.lon <- as.numeric(blocks.les$cross1.lon)
blocks.les$cross2.lat <- as.numeric(blocks.les$cross2.lat)
blocks.les$cross2.lon <- as.numeric(blocks.les$cross2.lon)

oldSame <- blocks.les[blocks.les$id %in% same$id, ]
oldSame$int1 = paste(oldSame$street, "&", oldSame$cross1)
oldSame2 <- data.frame(int2 = paste(oldSame$street, "&", oldSame$cross2),
                      cross2.lat = oldSame$cross2.lat, 
                      cross2.lon = oldSame$cross2.lon)
g <-  ggmap(mymap) + 
      theme_nothing() +
      geom_point(aes(x = cross1.lon, y = cross1.lat), data = oldSame, color = "red", size = 5) +
      geom_text(aes(label = int1, x = cross1.lon, y = cross1.lat), data = oldSame, color = "blue",
                check_overlap  = TRUE) +
      geom_point(aes(x = cross2.lon, y = cross2.lat), data = oldSame2, color = "red",   size = 5) +
      geom_text(aes(label = int2, x = cross2.lon, y = cross2.lat), data = oldSame2, color = "blue",
                 check_overlap  = TRUE)
g
```

There's still some minute discrepancies, like the triangle of Grand/Clinton/E Broadway/
Samuel Dickstein. But overall the new table is more accurate than the one with rows having
the same cross street for both sides. 

Blocks are collapsed so both sides of the street fall into a single block. 

```{r}
blocks.agg <- blocks.les %>% group_by(street, cross1, cross1.lat, cross1.lon, cross2, 
                         cross2.lat, cross2.lon) %>% 
  summarize(start = min(start), end = max(end), trees = sum(count)) %>% 
  as.data.frame()
# Remove incomplete cases (which have a 0 count anyway)
blocks.agg <- blocks.agg[complete.cases(blocks.agg), ]
```

At this point it can be determined which blocks have the most trees. 

```{r}
head(blocks.agg[order(blocks.agg$trees, decreasing = TRUE), c(1, 2, 5, 10)], 10)
```

Not surprisingly, the long blocks between avenues have the highest count of trees, all of which 
are in the East Village.

```{r eastVill}
map.evill <- get_map(location = "40.7228765,-73.9805015", zoom = 17, maptype = "toner-lines")
blocks.agg$int1 <- paste(blocks.agg$street, "&", blocks.agg$cross1)
  
g <- ggmap(map.evill) + 
  geom_point(data = treeMap, aes(x = lon, y = lat, size = stumpDiam, color = stumpDiam),
             alpha = 0.35) +
  scale_size(range = c(5, 7)) +
  theme_nothing() +
  scale_color_continuous(low = "chartreuse", high = "chartreuse4") +
  geom_text(aes(label = int1, x = cross1.lon, y = cross1.lat), data = blocks.agg, color = "red",
                fontface = "bold", check_overlap  = TRUE)
blocks.agg$int1 <- NULL
g
```

"Which block is the most tree-lined" can be answered now, but that answer is biased. Longer blocks
have an advantage. A more accurate measure would be trees per meter (TPM). To determine the distance
of a block, the Great-circle distance is used with the Haversine formula.

```{r}
deg2rad <- function(deg) return(deg * pi / 180)

gcd <- function(long1, lat1, long2, lat2) {
  # Computes the great-circle distance between two points using the haversine formula
  #
  # Args:
  #   long1:  Longitude of first point in decimal degrees
  #   lat2:   Latitude of first point in decimal degrees
  #   long2:  Longitude of first point in decimal degrees
  #   lat2:   Latitude of second point in decimal degrees
  #
  # Returns:
  #   The distance in meters
  
  long1 <- deg2rad(long1) 
  lat1 <- deg2rad(lat1)
  long2 <- deg2rad(long2) 
  lat2 <- deg2rad(lat2)
  r <- 6371000
  long.delt <- (long2 - long1)
  lat.delt <- (lat2 - lat1)
  a <- sin(lat.delt / 2)^2 + cos(lat1) * cos(lat2) * sin(long.delt / 2)^2
  c <- 2 * atan2(sqrt(a), sqrt(1 - a))
  r * c
}
```

```{r}
blocks.agg$distance <- NULL
for (i in 1:nrow(blocks.agg)) {
  d <- gcd(blocks.agg[i, "cross1.lon"], 
           blocks.agg[i, "cross1.lat"], 
           blocks.agg[i, "cross2.lon"], 
           blocks.agg[i, "cross2.lat"])
  blocks.agg[i, "distance"] <- d 
}

blocks.agg$tpm <- blocks.agg$trees / blocks.agg$distance
head(blocks.agg[order(blocks.agg$tpm, decreasing = TRUE), c(1, 2, 5, 10:12)], 10)
```

This gives an answer, but how valid is it? The preliminary checks above looked reasonably good, 
but can the block of Grand between Clinton and Suffolk really have so many more trees per meter than
any other block? A closer inspection can isolate the trees mapped to just that block. 

```{r sewardPark}
blocks.les[blocks.les$street == "Grand St" & blocks.les$cross1 == "Clinton St",]
treeMap[treeMap$blockId == 35, ]
map.sewardPark <- get_map(location = "40.71597,-73.9889687", zoom = 17, maptype = "toner-lines")
ggmap(map.sewardPark) + 
   geom_point(data = treeMap[treeMap$blockId == 35,],
              aes(x = lon, y = lat, size = stumpDiam), alpha = 0.35, color = "chartreuse") +
   scale_size(range = c(5, 7)) +
   theme_nothing()
```

The majority of the trees actually lie along Clinton, although they were coded at 410 Grand. 
This highlights the roughness of geocoding coordinates to addresses, and the limitations of the 
tools available. The manual work using the above procedures make it unfeasible to report over the 
entire city, which includes more than 300,000 trees. 
At this point, there's enough certainty to say that the blocks from this sample data set 
with the most trees per meter (and trees in general) are between Avenues C and D. Here is the 
complete map of all the trees from the original request.

```{r lesMap}
g <- ggmap(mymap) + 
  geom_point(data = treeMap,
             aes(x = lon, y = lat, size = stumpDiam, color = stumpDiam), 
             alpha = 0.35) +
  scale_size(range = c(5, 7)) +
  theme_nothing() +
  scale_color_continuous(low = "chartreuse", high = "chartreuse4")
g
```

[^1]: **New York City Street Tree Map Beta**  
  Interactive map to view details from a city-wide to a single tree level. No official API  
  https://tree-map.nycgovparks.org/points/\<SW lat>\/\<SW lng\>/\<NE lat\>/\<NE lng\>/undefined/
  \<start tree\>/\<end tree\>  
  View details of a specific tree, including closest address:  
  https://tree-map.nycgovparks.org/tree/full/\<tree-id from points json\>  
  Browser-based map:  
  https://tree-map.nycgovparks.org/ &nbsp;&nbsp;&nbsp;
  
[^2]: **ArcGIS REST API: World Geocoding Service**  
  ReverseGeocode returns the closest address to a given lat/lng location  
  http://developers.arcgis.com/rest/geocode/api-reference/geocoding-reverse-geocode.htm  
  The Find operation geocodes input and can handle intersections  
  https://developers.arcgis.com/rest/geocode/api-reference/geocoding-find.htm  &nbsp;&nbsp;&nbsp;
  
[^3]: **US Census Geocoding Services**  
  Geocodes addresses and also returns the range of addresses on the block of input
  http://geocoding.geo.census.gov/geocoder/Geocoding_Services_API.html  
  Also available as browser-based app: http://geocoding.geo.census.gov/geocoder/locations/address 
  &nbsp;&nbsp;&nbsp;